---
title: "Tidyverse"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r, include = F}

library(tidyverse)

```


# Tidyverse
[Tidyverse](https://www.tidyverse.org/packages/) is an R package that allow effective data manipulation, visualization and more. In this section, enter the following script upfront (and every time you open a new R session) so you can use functions in `tidyverse` package:

```{r, eval = F}

install.packages("tidyverse")
library(tidyverse)

```


## Data transformation

Data transformation is a critical component of data analysis. You can use spreadsheets or similar to manage your data. However, this form of data manipulation/transformation should be limited to the process of creating the primary dataset. There are several reasons for this. **First and foremost, a spreadsheet does NOT keep records of your edits.** If you make any mistake during the process of data transformation, those will be unnoticed. **Second, a spreadsheet is not well suited to data transformation.** You may need to iterate the same transformation multiple times, and you will make an error sooner or later.

### Filter rows

### Select columns

### Add new variables

### Grouping

### Drop NAs

### 
