---
title: "Basics"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Data Types and Structures

R has 6 basic **data types**.

- character: `"aquatic"`, `"ecology"` (no order)
- factor: similar to character, but has *levels* (alphabetically ordered by default)
- numeric: `20.0` , `15.5`
- integer: `3`, `7`
- logical: `TRUE` , `FALSE`
- complex: `1+2i` (complex numbers with real and imaginary parts)

These elements form one of the following **data structures**.  

- vector: a series of elements. ***A single data type is allowed in a single vector***
- matrix: elements organized into rows and columns. ***A single data type is allowed in a single matrix***
- data frame: looks similar to a matrix, but allows ***different data types in different columns***

# Vector
## Vector - Create Vector

Below are examples of atomic character vectors, numeric vectors, integer vectors, etc. There are many ways to create vector data. The following examples use `c()`, `:`, `seq()`, `rep()`:

```{r}

#ex.1a manually create a vector using c()
X <- c(1,3,4,8)
X

#ex.1b character
X <- c("a", "b", "c")
X

#ex.1c logical
X <- c(TRUE, FALSE, FALSE)
X

#ex.2 sequence of numbers
X <- 1:5
X

#ex.3a replicate same numbers or characters
X <- rep(2, 5) # replicate 2 five times
X

#ex.3b replicate same numbers or characters
X <- rep("a", 5) # replicate "a" five times
X

#ex.4a use seq() function
X <- seq(1, 5, by = 1)
X

#ex.4b use seq() function
X <- seq(1, 5, by = 0.1)
X

```

## Vector - Check Features

R provides many functions to examine features of vectors and other objects, for example:

- `class()` - what kind of object (data structure) is it (high-level)?
- `typeof()` - what is the object’s data type (low-level)?
- `attributes()` - does it have any metadata?
- `length()` - how long is it? What about two dimensional objects?
- `sum()` - what is the summed number of the data?
- `mean()` - what is the average number of the data?

### Numeric Vector

```{r}

X <- c(1.2, 3.1, 4.0, 8.2)
X
class(X)
typeof(X)
length(X)
sum(X)
mean(X)

```

### Character Vector

```{r}

Y <- c("a", "b", "c")
class(Y)
length(Y)

```
  
## Vector - Access
### Element ID

Use brackets `[]` when accessing specific elements in an object.
For example, if you want to access element #2 in the vector `X`, you may specify as `X[2]`:

```{r}

X <- c(2,2,3,2,5)
X[2] # access element #2
X[c(2,4)] # access elements #2 and 4
X[2:4] # access elements #2-4

```

### Equation

R provides many ways to access elements that suffice specific conditions. You can use mathematical symbols to specify what you need, for example:

- `==` equal
- `>` larger than
- `>=` equal & larger than
- `<` smaller than
- `<=` equal & smaller than
- `which()` a function that returns element # that suffices the specified condition

The following examples return a logical vector indicating whether each element in X suffices the specified condition:

```{r}

# creating a vector
X <- c(2,2,3,2,5)

# ex.1a equal
X == 2

# ex.1b larger than
X > 2 

```

You can access elements that suffice the specified condition using brackets, for example:

```{r}

# ex.2a equal
X[X == 2]
# ex.2b larger than
X[X > 2]

```

Using `which()`, you can see which elements (i.e., #) matches what you need:

```{r}

# ex.3a equal
which(X == 2) # returns which elements are equal to 2
# ex.3b larger than
which(X > 2)

```

# Matrix
## Matrix - Create Matrix

Matrix is a set of elements (*single data type*) that are organized into rows and columns:

```{r}

#ex.1 cbind: combine objects by column
X <- cbind(c(1,2,3), c(4,5,6))
X

#ex.2 rbind: combine objects by row
X <- rbind(c(1,2,3), c(4,5,6))
X

#ex.3 matrix: specify elements and the number of rows (nrow) and columns (ncol)
X <- matrix(1:9, nrow = 3, ncol = 3)
X

```

## Matrix - Check Features

R provides many functions to examine features of matrix data, for example:

- `class()` what kind of object (data structure) is it (high-level)?
- `typeof()` what is the object’s data type (low-level)?
- `attributes()` does it have any metadata?
- `dim()` how long are rows and columns?
- `rowSums()` what is the summed number of the data for each row?
- `colSums()` what is the summed number of the data for each column?

### Integer Matrix

```{r}

X <- matrix(1:9, nrow = 3, ncol = 3)
X
class(X)
typeof(X)
dim(X)

```

### Character Matrix

```{r}

Y <- matrix(c("a","b", "c", "d", "e", "f"), nrow = 3, ncol = 2)
Y
class(Y)
typeof(Y)
dim(Y)

```

## Matrix - Access

When accessing matrix elements, you need to pick row(s) and/or column(s), for example:

```{r}

X <- matrix(1:9, nrow = 3, ncol = 3)
X
X[2,3] # access an element in row #2 and colum #3
X[2,] # access elements in row #2
X[c(2,3),] # access elements in rows #2 and 3
X[,c(2,3)] # access elements in columns #2 and 3

```

You can assess each element with mathematical expressions just like vectors:

```{r}

X == 2 # equal
X > 2 # larger than

```

However, care must be taken when accessing elements, as it will be automatically converted to vector data:

```{r}

X[X == 2] # equal
X[X > 2] # larger than

```

`which()` needs an additional argument to return both row and column #:

```{r}

which(X == 2, arr.ind = TRUE)
which(X > 2, arr.ind = TRUE)

```

# Data Frame
## Data Frame - Basics

Data frame is a set of elements that are organized into rows and columns, but differ in several ways.

- it allows *multiple data types* in different columns
- each column has its *name*
- access columns by name (using `$`)

***Data frame is the most common data structure when manipulating ecological data***. A data set loaded from a spread sheet (we will address this later) will be automatically recognized as a data frame. Here is an example:

```{r}

# Create data frame
X <- c("Pristine", "Pristine", "Disturbed", "Disturbed", "Pristine") # Lake type
Y <- c(1.2, 2.2, 10.9, 50.0, 3.0) # TSS: total suspended solids (mg/L)
dat <- data.frame(LakeType = X, TSS = Y) # X is named as "LakeType" while Y is named as "TSS"
dat
colnames(dat) # call column names
dat$LakeType # access LakeType
dat$TSS # access TSS
class(dat) # check feature
class(dat$LakeType) # check feature of LakeType
class(dat$TSS) # check feature of TSS

```

You can access elements like a matrix as well:

```{r}

dat[,1] # access column #1
dat[1,] # access row #1
dat[c(2,4),] # access row #2 and 4

```

## Data Frame - Advanced

In a data frame, you can pick a subset of data. For example, you might want to pick data only from pristine lakes. You can do this by applying the methods in the vector/matrix section:

```{r}

dat1 <- dat[dat$LakeType == "Pristine",] # access Pristine lake type
dat1
dat2 <- dat[dat$LakeType == "Disturbed",] # access Disturbed lake type
dat2

```

This allows you to calculate mean TSS for each lake type separately:

```{r}

mean(dat1$TSS)
mean(dat2$TSS)

```

R provides many useful functions for summarizing data. Especially, `apply` family functions are very useful to summarize your data. Here, we use `tapply()` function as an example. We use a vector of values that we want to summarize in the first argument, and put "INDEX" variable indicating group names (in this case, `"Pristine"`" or `"Disturbed"`).

```{r}

tapply(X = dat$TSS, INDEX = dat$LakeType, FUN = mean)

```

This matches what we calculated above using `dat1` and `dat2`, but with much less efforts.


# Exercise
## Basics
### Vector

- Create three numeric vectors with length 3, 6 and 20, respectively. Each vector must be created using different functions in R.
- Create three character vectors with length 3, 6 and 20, respectively. Each vector must be created using different functions in R.

### Matrix
- Create a numeric matrix with 4 rows and 4 columns. Each column must contain identical elements.
- Create a numeric matrix with 4 rows and 4 columns. Each row must contain identical elements.
- Create a character matrix with 4 rows and 4 columns. Each column must contain identical elements.
- Create a character matrix with 4 rows and 4 columns. Each row must contain identical elements.
- Check the data structure (high-level) of each matrix.

### Dataframe
- Create a dataframe with 3 variables (name variables as `X`, `Y` and `Z`. `X` must be `character` while `Y` and `Z` must be `numeric`.
- Check the data structure (higher-level) of `X`, `Y` and `Z`

## Advanced
### Vector

Copy the following script to your R script:

```{r, eval = FALSE}

set.seed(1)
x <- rnorm(100)

```

- Identify element IDs of `x` that are greater than `2.0`
- Identify element values of `x` that are greater than `2.0`

### Matrix

Copy the following script to your R script:

```{r, eval = FALSE}

set.seed(1)
X <- matrix(rnorm(100), nrow = 10, ncol = 10)

```

- Identify element IDs of `X` that are greater than `2.0` (**specify row and column IDs**)
- Identify element values of `X` that are greater than `2.0` and caluculate the mean.

### Data Frame

Copy the following script to your R script:

```{r, eval = FALSE}

set.seed(1)
x <- rnorm(100, mean = 10, sd = 3)
y <- rpois(100, lambda = 10)
z <- rep(c("VA", "NC"), 50)
dat <- data.frame(temperature = x, abundance = y, state = z)

```

- Calculate means of `temperature` and `abundance` for states `VA` and `NC` separately
